---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# foRgotten

<!-- badges: start -->

<!-- badges: end -->

## Description

The foRgotten library extends the theory of forgotten effects with the aggregation of multiple key informants for complete graphs and chain bipartite graphs. Provides analysis tools for direct effects and forgotten effects.

The package allows for:

-   Calculation of the average incidence by edges for direct effects.
-   Calculation of the average incidence per row and column for direct effects.
-   Calculation of the median betweenness centrality per node for direct effects.
-   Calculation of the forgotten effects.
-   Use of complete graphs and chain bipartite graphs.

## Authors

**Elliott Jamil Mardones Arias** School of Computer Science Universidad Católica de Temuco Rudecindo Ortega 02351 Temuco, Chile [elliott.mardones\@uct.cl](mailto:elliott.mardones@uct.cl)

**Julio Rojas-Mora** Department of Computer Science Universidad Católica de Temuco Rudecindo Ortega 02351 Temuco, Chile and Centro de Políticas Públicas Universidad Católica de Temuco Temuco, Chile [julio.rojas\@uct.cl](mailto:julio.rojas@uct.cl)

## Installation

You can install the stable version of foRgotten from CRAN with:

``` r
# install.packages(“foRgotten”)
```

and the development version from GitHub with:

```{r}
#install.packages(“devtools”)
#devtools::install_github("ElliottMardones/foRgotten")
```

## Usage

```{r example}
library(foRgotten)
```

## 

## Functions

The package provides four functions:

```{r echo=TRUE}
?directEffects
```

Computes the mean incidence, left one-sided confidence interval, and p-value with multiple key informants for complete graphs and chained bipartite graphs. For more details, see help (de.sq).

```{r echo=TRUE}
?bootMargin
```

Computes the mean incidence for each cause and each effect, confidence intervals, and p-value with multiple key informants for complete graphs and chain bipartite graphs. For more details, see help(bootMargin).

```{r echo=TRUE}
?centrality
```

Performs the computation of median betweenness centrality with multiple key informants for complete graphs and chain bipartite graphs. For more details, see help(centrality).

```{r echo=TRUE}
?FE
```

Performs the forgotten effects calculation proposed by Kaufman and Gil-Aluja (1988) with multiple experts. The parameters allow you to specify the significant degree of truth and the order of incidence that is required to be calculated for chained bipartite graphs For more details, see help(FE).

## DataSet

The library provides 3 three-dimensional incidence matrices which are called `CC`, `CE` and `EE`. The data are those used in the study "Application of the Forgotten Effects Theory For Assessing the Public Policy on Air Pollution Of the Commune of Valdivia, Chile" developed by Manna, E. M et al (2018).

The data consists of 16 incentives, 4 behaviors and 10 key informants, where each of the key informants presented the data with a minimum and maximum value for each incident. The description of the data can be seen in Tables 1 and 2 of Manna, E. M et al (2018).

The book store presents the data with the average between the minimum and maximum value for each incidence, A being the equivalent to incentives and B to behaviors. For more details of the data you can consult:

```{r cars}
help(CC)
help(CE)
help(EE)
```

## Examples

### **directEffects()**

The `directEffects()` function calculates the mean incidence, left-sided confidence interval, and p-value for complete and chained bipartite graphs using multiple key informants. This function performs a t-test with left one-sided contrast via bootstrap BCa.

The function returns a list object with the subset of data `$DirectEffects` that includes the following information:

-   From: Origin of the incidence.
-   To: Destination of the incidence.
-   Mean: Average incidence.
-   UCI: Upper confidence interval.
-   p.value: Calculated p-value.

### Parameters

-   **CC**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix. By default, `CC = NULL`.
-   **CE**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix (for complete graphs) or a rectangular matrix (for chained bipartite graphs).
-   **EE**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix. By default, `EE = NULL`.
-   **thr**: Defines the degree of truth in which incidence is considered significant within the range [0,1]. By default, `thr = 0.5`.
-   **conf.level**: Defines the confidence level. By default, `conf.level = 0.95`.
-   **reps**: Defines the number of bootstrap replicates. By default, `reps = 10000`.
-   **delete**: Removes the non-significant results from the `$DirectEffects` set and returns the entered three-dimensional incidence arrays by assigning zeros to the edges whose incidences are significantly lower than thr at the p-value set in the `conf.level` parameter. By default, `delete = FALSE`.

#### Example: Complete Graphs

For complete graphs, only the `CE` parameter is used. For example:

```{r}
result <- directEffects( CE = CC, thr = 0.5, reps = 1000)
```

The results obtained correspond to 240 first-order incidences. The first six values are:

```{r}
head(result$DirectEffects)
```

Any result that contains an NA value in the UCI and p.value fields indicates that all incidences are equal or that the value is unique. Therefore, no bootstrapping is performed.

The parameter `delete = TRUE` removes non-significant incidents. For example:

```{r}
result <- directEffects( CE = CC, thr = 0.5, reps = 1000, delete = TRUE)

```

The number of significant incidences decreased from 240 to 217. For `delete = TRUE`, the function returns the entered three-dimensional incidence array but assigns zero to the non-significant edges.

#### Example: Chained bipartite graphs

The `CC`, `CE`, and `EE` parameters are used for chained bipartite graphs. The other parameters maintain their functionality except for the `delete` parameter, which now returns the three entered three-dimensional incidence arrays but assigns zero to the non-significant edges.

```{r}
result <- directEffects( CC = CC, CE = CE, EE = EE, thr = 0.5, reps = 1000, delete = TRUE)
names(result)

```

### bootMargin()

The `bootMargin()` function calculates the mean incidence of each cause and each effect, confidence intervals, and p-value for complete and chained bipartite graphs using multiple key informants. This function performs a t-test with bilateral contrast via bootstrap BCa.

The function returns a list object with the subset of data `$byCause` and `$byEffect`, which includes the following information:

-   Var: Name of the variable.

-   Mean: Average incidence.

-   LCI: Lower confidence interval.

-   UCI: Upper confidence interval.

-   p.value: calculated p-value.

### Parameters

-   **CC**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix. By default, `CC = NULL`.
-   **CE**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix (for complete graphs) or a rectangular matrix (for chained bipartite graphs).
-   **EE**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix. By default, `EE = NULL`.
-   **no.zeros:** For `no-zeros = TRUE`, the function omits zeros in the calculations. By default, `no-zeros = TRUE`.
-   **thr.cause**: Defines the degree of truth in which incidence is considered significant within the range [0,1]. By default, `thr = 0.5`.
-   **thr.effect**: Defines the degree of truth in which incidence is considered significant within the range [0,1]. By default, `thr = 0.5`.
-   **conf.level**: Defines the confidence level. By default, `conf.level = 0.95`.
-   **reps**: Defines the number of bootstrap replicates. By default, `reps = 10000`.
-   **delete**: Removes the non-significant results from the `$DirectEffects` set and returns the entered three-dimensional incidence arrays by assigning zeros to the edges whose incidences are significantly lower than `thr` at the p-value set in the `conf.level` parameter. By default, `delete = FALSE`.
-   **plot:** Generates a Dependence-Influence plot with the data from `$byCause` and `$byEffect`. The "Dependence" associated with `$byEffect` is on the X-axis, and the "Influence" associated with `$byCause` is on the Y-axis.

#### Example: Complete Graphs

For complete graphs, only the `CE` parameter is used. For example:

```{r}
result <- bootMargin( CE = CC, thr.cause = 0.5, thr.effect = 0.5, reps = 1000)
```

The results are:

```{r}
result$byCause
```

```{r}
result$byEffect
```

The parameter `delete = TRUE` removes non-significant incidents. For example:

```{r}
result <- bootMargin( CE = CC, thr.cause = 0.5, thr.effect = 0.5, reps = 1000, delete = TRUE)
```

The I14 variable was removed due to not being significant.

```{r}
result$byCause
```

```{r}
result$byEffect
```

The `plot = TRUE` parameter allows for the generation of the Dependence-Influence plot.

```{r}
result <- bootMargin( CE = CC, thr.cause = 0.5, thr.effect = 0.5, reps = 1000, delete = TRUE, plot = TRUE)
result$plot
```

#### Example: Chained bipartite graphs

The `CC`, `CE`, and `EE` parameters are used for chained bipartite graphs. The other parameters maintain their functionality except for the delete parameter, which now returns the three entered three-dimensional incidence arrays but assigns zero to the non-significant edges.

```{r}
result <- bootMargin( CC = CC, CE = CE, EE = EE, thr.cause = 0.5, thr.effect = 0.5, reps = 1000, delete = TRUE)
names(result)
```

### centralitry()

The `centrality()` function calculates the median betweenness centrality, confidence intervals, and the selected method for calculating the centrality distribution for complete and chained bipartite graphs using multiple key informants.

The function returns an object of type data.frame that contains the following components:

-   Var: Name of the variable.
-   Median: Median calculated.
-   LCI: Lower Confidence Interval.
-   UCI: Upper Confidence Interval.
-   Method: Statistical method used associated with the model parameter.

### Parameters

-   **CC**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix. By default, `CC = NULL`.
-   **CE**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix (for complete graphs) or a rectangular matrix (for chained bipartite graphs).
-   **EE**: It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix. By default, `EE = NULL`.
-   **model:** Allows you to determine to which heavy-tailed distribution the entered variables correspond.
-   **conf.level**: Defines the confidence level. By default, `conf.level = 0.95`.
-   **reps**: Defines the number of bootstrap replicates. By default, `reps = 10000`.
-   **parallel:** Establece el tipo de operación en paralelo requerida. Las opciones son "multinúcleo", "snow" y "no". Por defecto `paralelo = "no"`.
-   **ncpus:** Define el número de núcleos a utilizar. Por defecto `ncpus = 1`.

#### Example: Complete graphs

```{r}
result <- centrality(CE = CC, model = "median", reps = 300)
result
```

#### Example: Chained bipartite graphs

```{r}
result <- centrality(CC = CC, CE = CE, EE = EE, model = "median", reps = 300)
result
```

### FE()

The `FE()` function calculates the forgotten effects, the frequency of their occurrence, the mean incidence, the confidence intervals, and the standard error for complete and chained bipartite graphs using multiple key informants. This function implements bootstrap BCa.

The function returns two list-type objects. The first is `$boot`, which contains the following information:

-   From: Indicates the origin of the forgotten effects relationships.
-   Through_x: Dynamic field that represents the intermediate relationships of the forgotten effects. For example, for order n there will be “though_1” up to “though\_ \<n-1\>”.
-   To: Indicates the end of the forgotten effects relationships.
-   Count: Number of times the forgotten effect was repeated.
-   Mean: Mean effect of the forgotten effect
-   LCI: Lower Confidence Intervals.
-   UCI: Upper Confidence Intervals.
-   SE: Standard error.

And the second is `$byExpert`, which contains the following information:

-   From: Indicates the origin of the forgotten effects relationships.
-   Through_x: Dynamic field that represents the intermediate relationships of the forgotten effects. For example, for order n there will be “though_1” up to “though\_ \<n-1\>”.
-   To: Indicates the end of the forgotten effects relationships.
-   Count: Number of times the forgotten effect was repeated.
-   Expert_x: Dynamic field that represents each of the entered experts.

### 
Parameters

-   **CC:** It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix. By default, `CC = NULL`.
-   **CE:** It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix (for complete graphs) or a rectangular matrix (for chained bipartite graphs).
-   **EE:** It allows for entering a three-dimensional incidence array, with each submatrix along the z-axis being a square incidence matrix. By default, `EE = NULL`.
-   **thr:** Defines the degree of truth in which incidence is considered significant within the range [0,1]. By default, `thr = 0.5`.
-   **maxOrder:** Defines the limit of forgotten effects to calculate (if they exist). By default, `maxOrder = 2`.
-   **reps:** Defines the number of bootstrap replicates. By default, `reps = 10000`.
-   **parallel:** Sets the type of parallel operation required. The options are "multicore", "snow", and "no". By default, `parallel = "no"`**.**
-   **ncpus:** Defines the number of cores to use. By default, `ncpus = 1`.

#### Example: Complete graphs

```{r}
result <- FE(CE = CC, thr = 0.5, maxOrder = 3, reps = 1000)
```

#### Example: Chained bipartite graphs

```{r}
result <- FE(CC = CC, CE = CE, EE = EE, thr = 0.5, maxOrder = 4, reps = 1000)
```

### 

## References

1.  Kaufmann, A., & Gil Aluja, J. (1988). Modelos para la Investigación de efectos olvidados, Milladoiro. Santiago de Compostela, España.

2.  Manna, E. M., Rojas-Mora, J., & Mondaca-Marino, C. (2018). Application of the Forgotten Effects Theory for Assessing the Public Policy on Air Pollution of the Commune of Valdivia, Chile. In From Science to Society (pp. 61-72). Springer, Cham.

3.  Freeman, L.C. (1979). Centrality in Social Networks I: Conceptual Clarification. Social Networks, 1, 215-239.

4.  Ulrik Brandes, A Faster Algorithm for Betweenness Centrality. Journal of Mathematical Sociology 25(2):163-177, 2001.

5.  Canty A, Ripley BD (2021). boot: Bootstrap R (S-Plus) Functions. R package version 1.3-28.

6.  Davison AC, Hinkley DV (1997). Bootstrap Methods and Their Applications. Cambridge University Press, Cambridge. ISBN 0-521-57391-2, <http://statwww.epfl.ch/davison/BMA/>.

7.  Newman, M. E. (2005). Power laws, Pareto distributions and Zipf's law. Contemporary physics, 46(5), 323-351.

8.  Gillespie, C. S. (2014). Fitting heavy tailed distributions: the poweRlaw package. arXiv preprint arXiv:1407.3492.

9.  Kohl, M., & Kohl, M. M. (2020). Package ‘MKinfer’.
